---
title:  "Final Project"
subtitle: "INFO 523"
author: "Meghan Edgerton"
date: "2023-04-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r code1, echo = TRUE, eval=TRUE}
library(dplyr)
library(readxl)
library(rgl)
library(adamethods)
library(ggplot2)
library(outliers)
library(DMwR2)
library(Rlof)
library(knitr)

p_path <- "data/PassingStats.xlsx"
re_path <- "data/ReceivingStats.xlsx"
ru_path <- "data/RushingStats.xlsx"

# Datasets
passing <- read_excel(p_path)
receiving <- read_excel(re_path)
rushing <- read_excel(ru_path)

```
This example demonstrates the use of the Local Outlier Factor (LOF) Algorithm for Outlier Detection on college football player data from the 2022 NCAA season. Passing, receiving, and rushing data was downloaded from www.sports-reference.com. The purpose of this application is to find outliers within a big population of potential football recruits and mimics a scenario of a football coach wanting to know which quarterbacks, running backs, and wide receivers stand out from the rest of the draft class.

\newpage
```{r code2, echo = TRUE, eval=TRUE}
# Local Outlier Factor (Passing)

# Removing columns that aren't applicable to passing
p <- passing %>% select(`Passing Yards`, `Pass Completions`)

outlier.scores <- lofactor(p, k=10)

outliers <- order(outlier.scores, decreasing=T)[1:10]

p.outliers <- passing[outliers,]

best.qb <- arrange(p.outliers, desc(`Pass Completions`), desc(`Passing Yards`))

top5.qb <- head(best.qb, 5)
top5.qb.f <- top5.qb %>% select(Rank, Player, School, Conference,
                                `Pass Completions`,
                              `Pass Completion Percentage`, `Passing Yards`,
                              `Passing TDs`)

kable(top5.qb.f, caption = "Outliers Detected in 2022 Passing Data")

```
\newpage
```{r code3, echo = TRUE, eval=TRUE}
ggplot() +
  geom_point(data = passing, aes(x = `Pass Completions`,
                          y = `Passing Yards`)) +
  geom_point(data = top5.qb, aes(x = `Pass Completions`,
                          y = `Passing Yards`), colour = "blue",
                          size = 4, shape = 18) +
  scale_y_continuous(n.breaks = 5, limits = c(1000,5000)) +
  theme_bw() 

```
**Figure 1** A scatterplot displaying the number of pass completions and the total passing yards for NCAA college quarterbacks from the 2022 season. The blue points represent the outliers detected using a local outlier factor (LOF) algorithm on pass completions and passing yards.

\newpage

```{r code4, echo = TRUE, eval=TRUE}
# Local Outlier Factor (Receiving)

# Removing columns that aren't applicable to receiving
re <- receiving %>% select(Receptions, `Receiving Yards`)

outlier.scores2 <- lofactor(re, k=10)

outliers2 <- order(outlier.scores2, decreasing=T)[1:10]

re.outliers <- receiving[outliers2,]

best.rec <- arrange(re.outliers, desc(`Receptions`), desc(`Receiving Yards`))

top5.rec <- head(best.rec, 5)
top5.rec.f <- top5.rec %>% select(Rank, Player,School, Conference, Receptions,
                                  `Receiving Yards`, `Receiving TDs`,
                                  `Avg Receiving Yards Per Reception`)

kable(top5.rec.f, caption = "Outliers Detected in 2022 Receiving Data")
```

\newpage
```{r code5, echo = TRUE, eval=TRUE}
ggplot() +
  geom_point(data = receiving, aes(x = `Receptions`,
                                 y = `Receiving Yards`)) +
  geom_point(data = top5.rec, aes(x = `Receptions`,
                                 y = `Receiving Yards`), colour = "red",
             size = 4, shape = 18) +
  scale_y_continuous(n.breaks = 5, limits = c(0,1500)) +
  theme_bw() 
```
**Figure 2** A scatterplot displaying the number of receptions and the total receiving yards for NCAA college wide receivers from the 2022 season. The red points represent the outliers detected using a local outlier factor (LOF) algorithm on total receptions and receiving yards.

\newpage
```{r code6, echo = TRUE, eval=TRUE}
# Local Outlier Factor (Rushing)

# Removing columns that aren't applicable to rushing
ru <- rushing %>% select(`Rush Attempts`, `Avg Rushing Yards Per Attempt`) %>%
  filter(`Avg Rushing Yards Per Attempt` > 1.0)

outlier.scores3 <- lofactor(ru, k=10)

outliers3 <- order(outlier.scores3, decreasing=T)[1:10]

ru.outliers <- rushing[outliers3,]

best.rush <- arrange(ru.outliers, desc(`Rushing Yards`), desc(`Rush Attempts`))

top5.rush <- head(best.rush, 5)
top5.rush.f <- top5.rush %>% select(Rank, Player, School, Conference, `Rush Attempts`,
                                    `Rushing Yards`, `Rushing TDs`, 
                                    `Avg Rushing Yards Per Attempt`)

kable(top5.rush.f, caption = "Outliers Detected in 2022 Rushing Data")

```

\newpage
```{r code7, echo = TRUE, eval=TRUE}
ggplot() +
  geom_point(data = rushing, aes(x = `Rush Attempts`,
                                   y = `Rushing Yards`)) +
  geom_point(data = top5.rush, aes(x = `Rush Attempts`,
                                  y = `Rushing Yards`), colour = "green",
             size = 4, shape = 18) +
  scale_y_continuous(n.breaks = 6, limits = c(0,1800)) +
  scale_x_continuous(n.breaks = 5, limits = c(50,350)) +
  theme_bw() 
```
**Figure 3** A scatterplot displaying the number of rushing attempts and the total rushing yards for NCAA college running backs from the 2022 season. The green points represent the outliers detected using a local outlier factor (LOF) algorithm on rushing attempts and average rushing yards per attempt. Average rushing yards per attempt was filtered to be greater than 1.0 to remove any additional players from the dataset that are not running backs.

\newpage
---
Conclusions & Future Work
---

Using a local outlier factor algorithm to detect outliers in a dataset of college football player statistics worked relatively well - the detected outliers seemed to have impressive metrics compared to the rest of the dataset. I found it interesting that the outliers ended up having a wide range of different "ranks" that were already in the dataset. For future work I think it would be more interesting to dive deeper into this and use a more complex model, like a neural network.  



